An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-08-18 co-authors/commented_versions/manuscript_edKGD.tex
Removed cache

\vspace{1em}
Corresponding author:

John L. Godlee

johngodlee@gmail.com

School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom

@@

{\LARGE{\textbf{Blinded Main Text File}}}

\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}

\normalsize{Running title: Woodland functionality in southern Africa}

@@

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

@@

% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.

@@

Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

@@

Compared to other forested ecosystems, dry tropical woodlands are highly structured by disturbance, mainly through fire, and in Africa notably by herbivory also \citep{}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth. In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation.

% COME BACK TO THIS: In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and it's environmental determinants.

Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.

A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. However, studied conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.

@@

% High levels of disturbance in southern African woodlands may moderate the observable BEFR through it's effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion. These fast-growing bushy species often proliferate in southern African woodlands when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}. Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
% KGD: This text is useful in demonstrating your biological understanding of the ecosystem. I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996). Or you could include this whole paragraph just after that one. The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may interact their effect on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1983}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e. moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).

\subsection*{Inter-related effects of tree species and structural diversity on AGB}

We found a consistent positive effect of tree species diversity on AGB across all models in this study. Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}. Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-structured  and poorly studied ecological system, southern African woodlands and savannas.

% In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists. We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.

Much of the total variation in AGB was driven by variation in stem density. Stem density also correlated with species diversity in our SEMs. It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB. The opposite is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species. We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level. This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.

% KGD: There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.

@@

We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}). A higher tree species diversity allows for a greater structural diversity of trees, i.e. greater variation in DBH and height. This may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light. Although we did not measure it here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e. wider variation in canopy shape and overall growth form; bushy understorey vs. emergent canopy, for example. Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}. Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this. In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity. While we did not have access to adequate data on disturbance history in our plots, previous studies have found that southern African woodlands with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.

We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}). At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.

@@

Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity. We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stem density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.

Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}. In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect on AGB. Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted. It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB. We deliberately excluded plots with very low stem density as they are not considered woodlands, but grassy savannas. It may be that by excluding the bottom end of this stem density continuum we prevented a relationship being observed between moisture and AGB/stem density. Additionally, due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
% KGD: NEED MORE ON THIS ABOVE, FOR CLARIFICATION OF WHAT YOU MEAN

@@

We expected a positive effect of soil fertility on AGB, but found no evidence of this in our models. We measured soil fertility using the observed variables of soil organic carbon content, sand particle content and Cation Exchange Capacity (CEC). In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2010, MORE}. 
% KGD: You have written little about soil fertility in this section, but I think you need to. There is at least one paper which could be taken to suggest a positive relationship between AGB and soil fertility (Coelho de Souza et al. 2019. NEE, see my publications).

>>>>

@@

\begin{center}
\LARGE{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}

\normalsize{Woodland functionality in southern Africa}
\end{center}

@@

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context and biome.

@@

\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, operating largely via an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

@@

% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
%KGD: I think you can delete that old abstract. New one is much snappier/better!

@@

Numerous studies have shown relationships between biodiversity and ecosystem function 
(e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. %These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
%KGD: I think it is good to reference those classic experimental studies in grasslands. So, even if not that exact sentence, something similar would be good.

@@

%Here you start justifying why this study is needed. The 'it hasn't been done here' justification is often poorly received by reviewers, and that is how your justification starts. However, the 'it depends on ecological context' justification works better. In terms of basic theory we might not expect a BEFR in disturbance-dominated systems. That is what I would lead with. The reason reviewers (and scientists in general) prefer that is because it gives greater generality to your study, allow conclusions to be extrapolated to other systems.  The text on dry tropical ecosystems (particularly the percentages) is useful and should come in here somewhere. It could come towards the end of this paragraph then, stand on its own in a separate paragraph or even find its way into the subsequent paragraph (probably not on that last one though...).
The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity, which is driven by interspecific competition \citep{Wright2017, Poorter2015, Sande2017a}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function. Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.

%You could also cite White for the first sentence in this paragraph to connect to classic literature
%Also, we keep needing to refer to 'southern African savannas and woodlands' or 'southern African woodlands and savannas' so perhaps an acronym is in order (SASW or SAWS). Connotation of the latter is a bit awkward even if it flows of the tongue better.
Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.

%I think this paragraph could be deleted. In a sense it is a more extended rehash of the end of preceding paragraph. As we aren't actually studying human impacts directly, it is a bit of a distraction. Some of this text if you want to keep it could work well in a conclusions section.
Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}. This has raised interest in how biodiversity influences ecosystem function in these ecosystems. Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}. While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.

A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. However, studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass %do you have a citation to support this statement? Would be good...
, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.

%This paragraph is a bit awkward and includes multiple themes. I think the first half could be reduced in spliced into the preceding paragraph. Meanwhile, the second half relates to your a priori predictions about whether you expect a BEFR in disturbance-dominated / stressful ecosystems. It seems like that text would be better placed above where you are discussing general theoretical reasons for why a BEFR might exist and under what environmental conditions you would expect to find one.
In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest. In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. Stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured in dry forested ecosystems.

@@

%\citet{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity. They suggest this effect is caused by increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity. We expect that this effect will occur in southern African woodlands as well. A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species which can effectively tolerate sub-optimal conditions. It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.
%KGD: I agree on deleting this paragraph. Some of the justification for looking at abundance evenness could actually be included (briefly) in methods.

%High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}. Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}. A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion. These fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}. Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
%KGD: This text is useful in demonstrating your biological understanding of the ecosystem. I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996). Or you could include this whole paragraph just after that one. The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
%I wasn't sure what exactly you meant to say with #3. I have edited to what I think it should say, but I may have missed the mark.

%To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
%KGD: I don't think it would hurt to mention that you are using SEMs as your statistical workhorse here. This is a matter of style though. Include if you want to.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability. i.e.  moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when the effects of tree species diversity, structural diversity and stem density were accounted for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).

\subsection*{Effect of tree species diversity on AGB}

% I would just write that you had a consistent positive effect across all models, as that is your general result with regard to this question.

We found a consistent positive effect of tree species diversity on AGB across all models in this study. Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}. Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-dominated and also poorly studied ecological system, southern African woodlands.

Much of the total variation in AGB was driven by variation in stem density. Stem density also correlated with species diversity in our SEMs. It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB. The opposite is also plausible, with increased stem density causing higher species richness through an increased probability of encountering new species. %Not sure what you mean by following sentence->
In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a causal effect of species diversity on stem density exists. We cannot decompose the relative effects of tree species richness and abundance evenness in our model %but hypothetically you could have, e.g. with multiple latent variables, so perhaps best not to mention this. You can make a statement about the relative importance of the two based on the model,no?
, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}. We suggest that an increase in tree species diversity through species richness and %italicise the 'and' here
evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light and resulting in greater total AGB at the plot level. This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.
%The text towards the end of this paragraph seems like it should go with the following section, no?
% There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.

@@

\subsection*{Structural diversity as a mechanism for the BEFR}
%the preceding section and this one are quite interrelated and each one is quite short. I wonder if it is worth combining them as a 'species diversity and interrelated factors driving AGB' or something like that with the interrelated factors being stem density and structural diversity

We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}). A higher tree species diversity allows for a greater diversity of tree functional forms %functional forms or greater variation in tree size? You have really only measured tree size, whereas functional form seems like it should be more than that, incl. canopy shape. I would be more circumspect on this. Conversely, the term structural diversity doesn't seem like as much of a stretch, i.e. its OK to me.
within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light. Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}. %worth mentioning the stem density result again here? It seems relevant.
In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity. While we did not have access to adequate data on disturbance history in our plots, previous studies have found that southern African woodlands with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.

@@

Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity. We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees and a greater stem density per unit area \citep{Kraaij2006, Shirima2015}.
%great resource availability can give greater stem density without increasing resource partitioning I would think. I suppose it depends how you define resource partitioning, but I am thinking of it as niche partitioning.

Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}. In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect on AGB. Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted. It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB. %preceding sentence confusing to me
We deliberately excluded plots with very low stem density as they are not considered woodlands, but grassy savannas. It may be that by excluding the bottom end of this stem density continuum we limited our ability to detect a relationship between moisture and AGB/stem density. Additionally, due to the high levels of adaptation of tree species to different levels of water availability in southern Africa, at the large scale we conducted our experiment, turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB. %need a bit more text to clarify what you mean here.

@@

%you have written nothing about soil fertility in this section, but I think you need to. There is at least one paper which could be taken to suggest a positive relationship between AGB and soil fertility (Coelho de Souza et al. 2019. NEE, see my publications).
